---
title: "Investigation Into Student Mental Health Issues"
author: "Jarred Petersen"
output:
  html_document:
    df_print: paged
---

## Mental Health issues by Gender, Age, GPA, College Year and Marital Status

After the COVID-19 pandemic, where people across the globe found themselves isolated away from loved ones in lock downs, the issue of Mental health has gained more coverage globally. Mental health and it's impact on people of all ages, ethnicity, race, gender, and sexual orientation is now discussed more openly on regular basis. The importance of Student mental health has also been a hot topic of debate. 
This Rmarkdown file will explore mental health issues in a group of 101 Students surveyed in 2020.
I will build visualizations to explore mental health issues by gender, age, Cumulative GPA, college year, course of study and marital status and report my findings. 

**Link to data set used:**
https://www.kaggle.com/datasets/shariful07/student-mental-health

```{r}
library(tidyverse)   #Load tidyverse package
library(readr)       #load readr package
library(ggplot2)     #load ggplot2 package
```


```{r}
mental_health <- read_csv("Student Mental health.csv")  #read in data as mental_health data frame
str(mental_health)                                      #View structure of the data frame
```

From viewing the data set, there is 11 variables and 101 observations. Before exploring the data frame, there is some cleaning of the data set that needs to be done. 
I will need to rename a number of the variables, recode some values, change character data to factors, and reshape the data frame to a longer format.

### Clean and Prepare Data for Visualization 

```{r}
mental_health1 <- mental_health %>%                 #create mental_health1 data frame
  rename('Gender' = 'Choose your gender',           #Rename variables - 'Gender" = "choose your gender"
         'Course'= 'What is your course?',          #Course = 'what is your course?'
         'Year' = 'Your current year of Study',     #Year = 'Your current year of study' 
         'CGPA' = 'What is your CGPA?',             #CGPA = what is your CGPA
         'Depression' = 'Do you have Depression?',  #Depression = 'Do you have Depression?' 
         'Anxiety' = 'Do you have Anxiety?',        #Anxiety = ' Do you have Anxiety?'
         'Panic Attacks' = 'Do you have Panic attack?', #Panic_Attacks = 'Do you have panic attack?'
         'Treatment' = 'Did you seek any specialist for a treatment?') %>%  #Treatment = 'Did you seek any specialist for a treatment?'
    mutate_at(c("Gender", "Year",                    #Mutate listed variables - Gender, year, 
              "Depression", "Anxiety", "Panic Attacks", "Treatment",#depression, anxiety, panic_attacks
              "Marital status", "Course", "CGPA"), as.factor) %>% #martial status, course, CGPA to factors
   mutate(Year = recode(Year, "year 1" = "Year 1",  #recode 'year 1' to 'Year 1'
                        "year 2" = "Year 2",        #recode 'year 2' to 'Year 2'
                        "year 3" = "Year 3",        #recode 'year 3' to 'Year 3'
                        "year 4" = "Year 4"))       #recode 'year 4' to 'Year 4'
``` 


```{r}
mental_health2 <- mental_health1 %>%                #mental_health2 from mental_health1 
  pivot_longer(8:10,                                #pivot longer columns 8:10
               names_to = "Type",                   #name of columns to 'Type'
               values_to = "Suffer") %>%            #Values to Suffer Variable
  mutate_at("Type", as.factor)                      #mutate Type to factor 
```

### Explore Data Set 

Now that I have cleaned and prepared the data frame, I can explore the data set.

```{r}
str(mental_health2)       #View structure of mental_health2
```

Viewing the structure of the data frame, I see that my cleaning has been successful. The data frame has 11 variables; Gender, Age, Course, Year(College year), CGPA(cumulative GPA), Marital Status, whether the student received treatment, type of mental illness and whether the student suffers from mental illness. 
There are 101 observations. 
Gender is a factor with 2 levels - Male and Female.  
Course has 48 levels. 
College Year has 4 levels - year 1, year 2, year 3, year 4. 
CGPA has is a factor with 5 levels.
Marital Status is a factor with 2 levels - yes and no.
Treatment is a factor with 2 levels - yes and no. 
Type is a factor with three levels - Depression, Anxiety, Panic Attacks.

The majority of the variables are categorical data in levels. 

There is also 303 observations as we have made the data frame longer. Now each student has 3 observations. 


```{r}
summary(mental_health2)    #view summary of mental_health2 data frame 
```

Above we have the summary statistics for each of the variables. These values are all x3 from the original data set because I made the data set longer so each student has 3 observations. 

```{r}
head(mental_health2)       #view first 6 rows of mental_health2 data frame
tail(mental_health2)       #view last 6 rows of mental_health2 data frame
```

Using the head() and tail() function allows us to few the first 6 and last 6 rows of the data set. This is another way to view the format of the data frame. 


### Visualization 1 - Mental Health Issues by Student Gender

Firstly, I will explore Mental health issues by student Gender. 

```{r}
table(mental_health1$Gender)  #View table of Gender variable 
```

As we can see from the original data set, 75 females answered the survey compare to 26 males. Therefore, the gender data is heavily skewed to the female side. This means that comparing mental health between the two genders will be difficult. However, I can compare the percentage of males/females suffering from the different types of illness. We can compare percentage of male students surveyed who suffer vs female students surveyed who suffer from mental illness. Due to the difference in female/male students surveyed, the data will still be skewed towards female.

```{r}
mental_health3 <- mental_health2 %>%                #mental_health3 from mental_health2 
  group_by(Gender, Type, Suffer) %>%                #Group by Gender, Type and Suffer
  summarise(count = n()) %>%                        #Count values
  mutate(Percentage = round(count/sum(count)*100))  #add percentage column by count/sum of count *100

mental_health3 %>%                                            #plotting mental_health3
  ggplot(aes(Suffer, Percentage,                              #plot Suffer on x, Percentage on Y
             fill = Suffer))+                                 #Fill color by suffer
  geom_bar(stat = "identity")+                                #plot bar chart
  facet_grid(Gender~Type)+                                    #Facet grid by Gender and Type
  scale_fill_manual("Suffer",                                 #scale fill "Suffer" manually
                    values = c("Yes" = "red",                 #Yes = Red
                               "No" = "lightblue"))+          #No = light blue
  labs(title = "Mental Health Issues By Gender",        #set labels - Main title
       x = "Suffering From Mental Illness",             #x = "Suffering From Mental Illness"
       y = "Percentage of Students")+                   #Y = "Percentage of Students"
  theme_minimal()+                                      #use Minimal theme
  theme(plot.title = element_text(size = 20,            #size of title text at 20, 
                                  face = "bold",        #make it bold,
                                  color = "black"),    #color of text = black. 
        axis.title.x = element_text(size = 14,          #set x axis title text to 14, 
                                    face = "bold",      #make it bold,
                                    color = "black"),  #color of text = black 
        axis.title.y = element_text(size = 14,          #set y axis title text to 14,
                                    face = "bold",      #make it bold,
                                    color = "black"),  #set text color to black
        strip.text.x = element_text(face = "bold",      #edit facet grid text on x - make bold
                                    color = "black"),  #color of text = black
        strip.text.y = element_text(face = "bold",      #edit facet grid text on y - make bold
                                    color = "black"),  #color of text = black
        legend.title = element_text(face = "bold",      #make legend title bold
                                    color = "black"))  #make text color black
```

1) Female - The most common mental illness in female students is depression. The percent of female students that suffer from panic attacks and anxiety is very similar, however, it appears that the percentage of female suffers of panic attacks is very slightly higher than anxiety. For all three mental illness types, there are less female students that suffer from the mental illness than those that do not suffer. 
2) Male - The most common mental illness in male students is anxiety, then panic attacks, and in 3rd depression. Similar to female students, there are more students that do not suffer from the mental illness than those that do suffer. 

Although there is fewer Students that suffer from mental illness than do suffer, the percentage of mental illness in both female and and male students is still significant. In female students, the percentage of students that suffer from mental illness is higher than 30% for all mental illness types. In male students, the percentage of students that suffer from mental illness is greater than 30% for anxiety and panic attacks.
If we compare male and female students, A higher percentage of male students suffer from anxiety than female students. A higher percentage of female students suffer from depression than male students. The percentage of male and female students suffering from panic attacks is very similar, however, the percentage of female students suffering from panic attacks is slightly higher than in male students.


### Visualization 2 - Mental Health Issues by College Year

Next, I will explore mental health issues by college year.

```{r}
mental_health3 <- mental_health2 %>%                #mental_health3 from mental_health2
  group_by(Year, Type, Suffer) %>%                  #group by Year, Type and Suffer
  summarise(Count = n()) %>%                        #count values using summarize 
  mutate(Percent = round(Count/sum(Count)*100))     #add percentage column by count/sum of count *100


mental_health3 %>%                                  #mental_health3 data frame
  ggplot(aes(Suffer, Percent,                       #plot suffer on x, Percentage on Y
             fill = Suffer))+                       #fill by Suffer
  geom_bar(stat = "identity")+                      #bar chart using "identity" stat 
  facet_grid(Year~Type)+                            #facet_grid by Year and type
  scale_fill_manual("Suffer",                       #scale fill color - Suffer variable
                    values = c("Yes" = "red",       #Yes = Red
                               "No" = "lightblue"))+#No = light blue
  labs(title = "Mental Health Issues By College Year",  #set labels - Main title
       x = "Suffering From Mental Illness",             #x = "Suffering From Mental Illness"
       y = "Percentage of Students")+                   #Y = "Percentage of Students"
  theme_minimal()+                                           #use Minimal theme
  theme(plot.title = element_text(size = 20,            #size of title text at 20, 
                                  face = "bold",        #make it bold,
                                  color = "black"),    #color of text = black. 
        axis.title.x = element_text(size = 14,          #set x axis title text to 14, 
                                    face = "bold",      #make it bold,
                                    color = "black"),  #color of text = black 
        axis.title.y = element_text(size = 14,          #set y axis title text to 14,
                                    face = "bold",      #make it bold,
                                    color = "black"),  #set text color to black
        strip.text.x = element_text(face = "bold",      #edit facet grid text on x - make bold
                                    color = "black"),  #color of text = black
        strip.text.y = element_text(face = "bold",      #edit facet grid text on y - make bold
                                    color = "black"),  #color of text = black
        legend.title = element_text(face = "bold",      #make legend title bold
                                    color = "black"))  #make text color black
```

Year 1) In Year 1, the percentage of students suffering of the types of mental illness are fairly even. 
Year 2) In year 2, the percentage of students suffering from anxiety and depression is higher than the percentage of students suffering from panic attacks.
Year 3) In year 3, a higher percentage of students suffer from depression and panic attacks than panic attacks.
Year 4) In year 4, the higher percentage of students suffer from anxiety than depression and panic attacks.

When we compare the 4 years, the percentage of students suffering of mental illness is much less in year 4 when compared to the other 3 years. Percentage of students suffering of mental illness in year 1, 2 and 3 is fairly significant. 
Anxiety is at its highest percentage in Year 2.
Depression is at its highest percentage in year 3.
Panic attacks is at its highest percentage in year 4.


### Visualization 3 - Mental Health Issues by Cumulative GPA Group

Next, I will explore Mental illness by cumulative GPA. 

```{r}
mental_health4 <- mental_health2 %>%        #Create mental_health4 data frame from mental_health2
  group_by(Year, CGPA, Type, Suffer) %>%          #group by CGPA Type, Suffer
  summarize(count = n())                    #Summarize count 


mental_health4 %>%                          #using mental_health4 data frame
  filter(Suffer == "Yes") %>%               #filter "Yes" suffer values 
  ggplot(aes(CGPA, count,                   #plot CGPA on x axis, count on Y
             fill = Type))+                 #fill colour by type
  geom_bar(stat = "identity")+              #plot bar chart using "identity" stat
  facet_grid(Type~Year)+                        #facet_grid by Type and Year
  scale_fill_manual("Type",                             #scale fill color manually 
                    values = c("Anxiety" = "grey",      #Anxiety = grey
                               "Depression" = "navy",   #Depression = Navy 
                               "Panic Attacks" = "darkred"))+ #Panic attacks = dark red
  labs(title = "Mental Health Issues by Cumulative GPA",#set labels - Main title
       x = "Cumulative GPA",                            #x = "Cumulative GPA"
       y = "Number of Students")+                       #Y = "Number of Students"
  theme_minimal()+                                      #use minimal theme
  theme(plot.title = element_text(size = 18,            #size of title text at 18, 
                                  face = "bold",        #make it bold,
                                  color = "black"),    #color of text = black. 
        axis.title.x = element_text(size = 14,          #set x axis title text to 14,
                                    face = "bold",      #make it bold,
                                    color = "black"),  #color of text = black 
        axis.text.x = element_text(angle = 45,          #edit x axis text , - angle of text set to 45
                                   hjust = 1),          #horizontal adjustment set to 1
        axis.title.y = element_text(size = 14,          #set y axis title text to 14,
                                    face = "bold",      #make it bold,
                                    color = "black"),  #set text color to black
        strip.text.x = element_text(face = "bold",      #edit facet grid text on x - make bold
                                    color = "black"),  #color of text = black
        strip.text.y = element_text(face = "bold",      #edit facet grid text on x - make bold
                                    color = "black"),  #color of text = black
        legend.title = element_text(face = "bold",      #make legend title bold
                                    color = "black"))  #make text color black
```

*Year 1*
In year 1, mental health issues arise from CGPA 2.50 - 4.00.

Anxiety) - Anxiety is at its highest in the highest GPA group of 3.50-4.00. Anxiety was at its lowest in the 2.50 -2.99 CGPA group. There was no anxiety cases below 2.49 group. This could be because there was no 1st year students with a CGPA of below 2.49. 
Depression) - Depression was at its highest in the 3.00 - 3.49 CGPA group. It was also significantly high in the 3.50 - 4.00 group. Similar to anxiety, there was no depression cases below a CGPA of 2.49.
Panic Attacks) - Panic attacks follow the same pattern as anxiety. Panic attacks were at its highest in CGPA of 3.50-4.00.

*Year 2*

Anxiety) - Anxiety only appears when CGPA is greater than 3.00. It was at its highest in the 3.50 - 4.00 CGPA group.
Depression) - Depression appears when GPA is greater than 2.50. It was at its highest in the 3.00 - 3.49 CGPA group.
Panic Attacks) - Panic attacks appear when CGPA is 2.0 - 4.0. However, there is only one student per CPGA group in 2.0 - 2.49, 2.5- 3.0, and 3.0 - 3.49  that suffers from panic attacks. It is at its highest in 3.50- 4.0 GPA. 

*Year 3*

In year 3, mental health issues only appear in CGPA groups of 3.00-3.49 and 3.50-4.0.

Anxiety) - Anxiety was at its highest in the 3.0 - 3.49 CGPA group.
Depression) - Depression was at its highest in the 3.0 - 3.49 CGPA group.
Panic Attacks)- Panic attacks was at it highest in the 2.5 - 4.0 CGPA group.

*Year 4*

Mental Health issues in year 4 only appear in the 3.0 - 3.49 group. 

The graph shows that the higher the CGPA, the more likely the student was to suffer from mental illness. 
Year 1 Students with a wider spread of CGPA were more likely to suffer from mental illness when compare with the other years, however, depression in year 2 appear across a more CGPA groups when compared with other years and mental illness. 
This chart confirms that mental illness issues are far less common in year 4 than year 1, 2 and 3. 


### Visualization 4 - Mental Health Issues by Age

The next chart will compare mental health issues by age. 

```{r}
mental_health2 %>%                                      #use mental health2 data frame
  filter(Suffer == "Yes") %>%                           #filter suffer by "yes"
  ggplot(aes(Age,                                       #plot Age on x axis
             fill = Type))+                             #Fill color by type
  geom_histogram(binwidth = 1)+                         #plot histogram + set bin width to 1
  facet_grid(~Type)+                                    #facet_grid by Type
  scale_fill_manual("Type",                             #scale fill color manually 
                    values = c("Anxiety" = "grey",      #Anxiety = grey
                               "Depression" = "navy",   #Depression = Navy 
                               "Panic Attacks" = "darkred"))+ #Panic attacks = dark red
  labs(title = "Mental Health Issues By Age",           #set labels - Main title
       x = "Age",                                       #x = "Age"
       y = "Number of Students")+                       #Y = "Number of Students"
  theme_minimal()+                                      #use minimal theme
  theme(plot.title = element_text(size = 20,            #size of title text at 20, 
                                  face = "bold",        #make it bold,
                                  color = "black"),    #color of text = black. 
        axis.title.x = element_text(size = 14,          #set x axis title text to 14, 
                                    face = "bold",      #make it bold,
                                    color = "black"),  #color of text = black 
        axis.title.y = element_text(size = 14,          #set y axis title text to 14,
                                    face = "bold",      #make it bold,
                                    color = "black"),  #set text color to black
        strip.text.x = element_text(face = "bold",      #edit facet grid text on x - make bold
                                    color = "black"),  #color of text = black
        legend.title = element_text(face = "bold",      #make legend title bold
                                    color = "black"))  #make text color black
```

Anxiety) - We see that anxiety is at its highest at the ages of 18, and 24. However, anxiety occurs more often in ages 18 - 21.
Depression) - Depression is at its highest at 18, and occurs more often in ages 18-20. Depression doesn't occur at age 21. 
Panic Attacks) - Panic Attacks occurs more at ages 18-19, and 24. Ages 21-22 do not suffer from panic attacks)

It appears that most mental illnesses appear at either end of the age range in college. Mental illnesses are most prevalent at age 18.


### Visualization 5 - Mental Health Issues by Martial Status

Next, I will compare mental illness by marital status.

```{r}
table(mental_health1$`Marital status`)    #View table of Martial Status variable
```

Similar to gender, the data is heavily distributed to one side. Far more unmarried students answered the survey than married students which makes sense given the age range of college students.
Therefore, I will compare what mental illnesses are most common in married and unmarried students as a percentage. 

```{r}
mental_health5 <- mental_health2 %>%                    #Mental_health5 from mental_health2
  group_by(`Marital status`, Type, Suffer) %>%          #group by Gender, marital status, type, suffer
  summarise(Count = n()) %>%                            #count values using summarize 
  mutate(Percent = round(Count/sum(Count)*100))     #add percentage column by count/sum of count *100
  


mental_health5 %>%                                     #Plot mental_health5 data frame
  filter(Suffer == "Yes") %>%                          #filter "Yes" values from suffer variable
  ggplot(aes(Type, Percent,                            #plot Type on x, count on y 
             fill = Type))+                            #fill color by Type    
  geom_bar(stat = "identity")+                         #plot bar chart using "identity" stat
  facet_grid(~`Marital status`)+                       #facet_grid by Marital status
   scale_fill_manual("Type",                             #scale fill color manually 
                    values = c("Anxiety" = "grey",      #Anxiety = grey
                               "Depression" = "navy",   #Depression = Navy 
                               "Panic Attacks" = "darkred"))+ #Panic attacks = dark red
  labs(title = "Mental Health Issues by Marital Status",#set labels - Main title
       x = "Illness",                                   #x = "Illness"
       y = "Percentage of Students")+                   #Y = "Percentage of Students"
  theme_minimal()+                                      #use minimal theme
  theme(plot.title = element_text(size = 18,            #size of title text at 18, 
                                  face = "bold",        #make it bold,
                                  color = "black"),    #color of text = black. 
        axis.title.x = element_text(size = 14,          #set x axis title text to 14,
                                    face = "bold",      #make it bold,
                                    color = "black"),  #color of text = black 
        axis.title.y = element_text(size = 14,          #set y axis title text to 14,
                                    face = "bold",      #make it bold,
                                    color = "black"),  #set text color to black
        strip.text.x = element_text(face = "bold",      #edit facet grid text on x - make bold
                                    color = "black"),  #color of text = black
        legend.title = element_text(face = "bold",      #make legend title bold
                                    color = "black"))  #make text color black
```

1) Unmarried - In Unmarried students, a higher percentage of unmarried students suffer from anxiety than the other two types of mental illness with 32% suffering from anxiety. 2nd highest percentage is Panic attacks at 27%, and 3rd is depression at 22%. A higher percentage of unmarried students do not suffer from mental illness than those that do. 

2)Married - Mental illness in married students is significantly high. Out of the 16 married students surveyed, 100% of them said that they suffered from depression, 62% suffer from panic attacks, and 44% suffer from anxiety. A higher percentage of married students students suffer from mental illness than those that do not. 

If we compare the unmarried and married, a much higher percentage of married students suffer from mental illness than unmarried students. 


### Visualization 6 - Mental Health Illness by Top 3 Courses

Next, I am going to visualize mental health issues by course of student. As there are 48 different courses in the original data frame, I will look at the top 3 courses.

```{r}
table(mental_health1$Course)  #view table of courses
```

From the above table, the top 3 courses i will be looking at is Engineering with 17 students,
BCS with 18 students, and BIT with 10 students.


```{r}
mental_health6 <- mental_health2 %>%                #create mental_heath7 from mental_health2
  filter(Course %in% c("BIT", "Engineering", "BCS")) %>%  #filter course = BIT, Engineering, BCS
  group_by(Course, Type, Suffer) %>%                #group by Course, Type, Suffer 
  summarise(Count = n()) %>%                        #summarize by adding count values
  mutate(Percent = round(Count/sum(Count)*100))     #add percentage column by count/sum of count *100


mental_health6 %>%                                      #Use mental_health6 data frame
  filter(Suffer == "Yes") %>%                           #Filter suffer by "Yes" values
  ggplot(aes(Type, Percent,                             #Plot Treatment on X, Percent on y
             fill = Type))+                             #Fill color by Type 
  geom_bar(stat ="identity")+                           #Plot bar chart with stat = Identity 
  facet_grid(~Course)+                                  #facet_grid by course
  scale_fill_manual("Type",                             #scale fill color manually 
                    values = c("Anxiety" = "grey",      #Anxiety = grey
                               "Depression" = "navy",   #Depression = Navy 
                               "Panic Attacks" = "darkred"))+ #Panic attacks = dark red
  labs(title = "Mental Illess By Degree Course",        #set labels - Main title
       x = "Type",                                      #x = "Did The Student Seek Treatment"
       y = "Percentage of Students")+                   #Y = "Percentage of Students"
  theme_minimal()+                                      #use minimal theme
  theme(plot.title = element_text(size = 20,            #size of title text at 20, 
                                  face = "bold",        #make it bold,
                                  color = "black"),    #color of text = black. 
        axis.title.x = element_text(size = 14,          #set x axis title text to 14, 
                                    face = "bold",      #make it bold,
                                    color = "black"),  #color of text = black 
        axis.title.y = element_text(size = 14,          #set y axis title text to 14,
                                    face = "bold",      #make it bold,
                                    color = "black"),  #set text color to black
        strip.text.x = element_text(face = "bold",      #edit facet grid text on x - make bold
                                    color = "black"),  #color of text = black
        axis.text.x = element_text(angle = 45,          #edit x axis text - angle set to 45
                                   hjust = 1),          #horizontal adjustment set to 1
        strip.text.y = element_text(face = "bold",      #edit facet grid text on y - make bold
                                    color = "black"),  #color of text = Black
        legend.title = element_text(face = "bold",      #make legend title bold
                                    color = "black"))  #make text color black
```

BCS) - In the BCS course, the most common mental illness is anxiety with 33% of the students suffering from it. 28% of BCS students suffer from depression and panic attacks
BIT) - In the BIT course, Students suffering with anxiety is significant with 80% of students suffering from anxiety. 50% of BIT students suffer from depression, and 40% of students suffer from panic attacks. 
Engineering) - A higher percentage of students in the engineering program suffer from depression, with 41% reporting they suffer. The 2nd highest percentage in panic attacks, at 29%. In third is anxiety with 24% of students suffering from anxiety.

When we compare the 3 courses, Mental health illness is far more prevalent in the BIT course when compared with the other two courses. Anxiety and depression appears to be the most common mental illness in the three courses.


### Visualization 7 - Do Students with Mental Health Issues Seek help?

As the previous graphs have shown, mental illness is a significant issue in students.

```{r}
mental_health7 <- mental_health2 %>%                #create mental_health6 from mental_health2
  filter(Suffer == "Yes") %>%                       #Filter "yes" Suffer values
  group_by(Gender, Type, Suffer, Treatment) %>%     #group values by Gender, Type, Suffer, Treatment
  summarise(count = n()) %>%                        #summarize values - count 
  mutate(Percent = round(count/sum(count)*100))     #add percentage column by count/sum of count *100


mental_health7 %>%                                      #Use mental_health6 data frame
  ggplot(aes(Treatment, Percent,                        #Plot Treatment on X, Percent on y
             fill = Treatment))+                        #Fill color by Treatment 
  geom_bar(stat ="identity")+                           #Plot bar chart with stat = Identity 
  facet_grid(Gender~Type)+                              #facet_grid by Gender and Type
   scale_fill_manual("Treatment",                       #scale fill "Treatment" manually
                    values = c("No" = "red",            #No = Red
                               "Yes" = "lightblue"))+   #Yes = light blue
  labs(title = "Students Who Sought Treatment",         #set labels - Main title
       x = "Did The Student Seek Treatment?",           #x = "Did The Student Seek Treatment"
       y = "Percentage of Students")+                   #Y = "Percentage of Students"
  theme_minimal()+                                      #use minimal theme
  theme(plot.title = element_text(size = 20,            #size of title text at 20, 
                                  face = "bold",        #make it bold,
                                  color = "black"),    #color of text = black. 
        axis.title.x = element_text(size = 14,          #set x axis title text to 14, 
                                    face = "bold",      #make it bold,
                                    color = "black"),  #color of text = black 
        axis.title.y = element_text(size = 14,          #set y axis title text to 14,
                                    face = "bold",      #make it bold,
                                    color = "black"),  #set text color to black
        strip.text.x = element_text(face = "bold",      #edit facet grid text on x - make bold
                                    color = "black"),  #color of text = black
        strip.text.y = element_text(face = "bold",      #edit facet grid text on y - make bold
                                    color = "black"),  #color of text = Black
        legend.title = element_text(face = "bold",      #make legend title bold
                                    color = "black"))  #make text color black
```

The above graph shows that out of the students that suffer of mental illness, a very small percentage actually seek help and treatment for it. 
Female students are more likely to seek treatment for depression and panic attacks than for anxiety. Male students are for more likely to seek treatment for depression, and no male students suffering from panic attacks sought treatment.
Female students a very slightly more likely to seek help than male students. 


### Conclusion 

To conclude, this data set shows that mental illness is a significant issue in students. 
A higher percentage of female students suffered from depression when compared anxiety and panic attacks, where as a higher percentage of male students suffered from anxiety then the other two types of mental illness, and depression was the lowest.

Far less students in year 4 of college suffer from mental illness when compared to years 1, 2 and 3. Year 2 and year 4 students are more likely to suffer from anxiety, and year 3 students are more likely to suffer from depression. The percentage of students suffering of the three types of mental illness in year 1 was fairly evenly spread.

The general trend of Mental Illness by CGPA is that as CGPA increases, number of mental illness cases increases. Mental illness tends to happen in the 2.5 - 4.0 region in years 1, 2, and 3. There a far fewer cases of mental illness in year 4, but these cases happen in the higher CGPA groups.

Mental illness was more prevalent in 18-19 year old and 24 year old. This is true for all three types of mental illness. 21 year old tended to not be suffering from mental illness. 

The percentage of mental illness cases in married students was significantly high. 100% percent of married students said they were suffering from depression, 62% suffer from panic attacks, and 44% suffer from anxiety. A higher percentage of married students students suffer from mental illness than those that do not. In unmarried students, there was a higher percentage of students that did not suffer from the mental illness type than those that did. Anxiety was the most common mental illness in unmarried students.

Out of the top 3 courses that students surveyed were studying, mental illness was most prevalent in BIT, where 80% of the students reported they suffered from anxiety. 50% of BIT students said they suffered from depression. In the BCS course, Anxiety was again the most common mental illness. In engineering, a higher percentage of students suffered from depression when compared to the other two mental illness types.

Although mental illness is very significant in students, a very small proportion of suffers actually sought treatment for help. None of the male students suffering from panic attacks sought treatment. Female students were slightly more likely to seek treatment for depression, however, a significantly larger percentage of female students suffering from the mental illness did not seek any help. 

This concludes my report and analysis on mental health issues in Students. 
